miniKanren with fair search strategies

### 2019-05-04 Paper/paper.tex
intro

\newcommand{\conde}{\texttt{cond$^e$} }
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}

@@

\conde is the flagship disjunction operator in miniKanren. A subtlety arises
when a \conde contains many disjunction clauses: not every clause has the equal
chance to contribute to the answers. As an example, consider the following
relation \texttt{repeato} and its invocation.
\clisting{Figures/repeato.rkt}
Next, consider the following disjuction of invoking \texttt{repeato} with
four different letters.
\clisting{Figures/example.rkt}
\conde intuitively corresponds to \texttt{or} in logic. An unsuspicious beginner
would expect each letter to contribute equally to the result, as following. 
\clisting{Figures/run-repeato-fair.rkt}
The \conde in TR2, however, generates a less expected result.
\clisting{Figures/run-repeato-idfs.rkt}

A subtlety arises when 
A subtlety in writing miniKanren programs with \conde expressions with many
clauses, such 
as relational interpreters, is that the order of \conde clauses sometimes 
affect the speed considerably. This phenomenon appears when running with the 
miniKanren in \citet{friedman_reasoned_2018}, one of the most well-understood 
implementation. This is because for all \conde expressions with more than two 
clauses, the clauses are given different ``search priority''. Left clauses are 
given higher priorities. This biased treatment causes 2 problems when a \conde 
expression has many clauses: the right-most clause can hardly contribute to 
query result; and the efficiency of the whole program might depend largely on 
the order of these clauses.

>>>>

\newcommand{\conde}{\texttt{cond$^e$} }
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}
\newcommand{\clisting}[1]{
\begin{center}
  \begin{tabular}{c}
	\lstinputlisting{#1}
    \end{tabular}
\end{center}
}
\newcommand{\DFSi}[0]{
  DFS$_\textrm{i}$
}

@@

\conde is the flagship disjunction operator in miniKanren. A subtlety arises
when a \conde contains many disjunction clauses: not every clause has the equal
chance to contribute to the result. As an example, consider the following
relation \texttt{repeato} and its invocation.
\clisting{Figures/repeato.rkt}
Next, consider the following disjuction of invoking \texttt{repeato} with
four different letters.
\clisting{Figures/example.rkt}
\conde intuitively corresponds to \texttt{or} in logic. And thus an unsuspicious
beginner would expect each letter to contribute equally to the result, as
following.
\clisting{Figures/run-repeato-fair.rkt}
The \conde in TR2, however, generates a less expected result.
\clisting{Figures/run-repeato-idfs.rkt}

\conde implements interleaving DFS(\DFSi), the cause of this unexpected result.
In a \conde, each clause takes a half of its received computational resources
and pass the other half to its following clauses, except for the last clause
that takes all resources it receives. In the example above, the letters
\texttt{c} and \texttt{d} are allocated with less resources than \texttt{a},
and thus \texttt{c} and \texttt{d} barely contribute to the result.

\DFSi is powerful for an expert. By carefully organizing the order of \conde
clauses, a miniKanren program can explore more ``interesting'' clauses than
those uninteresting ones, and thus use computational resources efficiently.
A little miniKanrener, however, may beg to differ--understanding
implementation details and fiddling with clauses order is not the first
priority of a beginner.

There is another reason that miniKanren could use more search strategies than
just \DFSi. In many applications, there does not exist one order that serves
for all purposes. For example, a relational dependent type checker contains
clauses for constructors that build data and clauses for eliminators that use
data. When this type checker is used to generate simple and shallow programs,
the clauses of constructors should be put in the front of \conde.
When performing proof search for complicated programs, the clauses of eliminator
should take the focus. With \DFSi, these two uses can not be efficient at the
same time. In fact, to make on use efficient, the other one must be drastically
slow.
